Testing the thermal buffer of mire groundwater

### 2020-05-07 doc/manuscript/ms.Rmd
intro drafted 2

Climate change is already affecting global biodiversity patterns (IPCC 2014). Among other trends, there has been an expansion of drylands (Huang, J., Yu, H., Guan, X. et al. Accelerated dryland expansion under climate change), an increase of disturbations in forests (Seidl, R., Thom, D., Kautz, M. et al. Forest disturbances under climate change. Nature Clim Change 7, 395–402 (2017). https://doi.org/10.1038/nclimate3303) and ocean acidification (Coral Reef Ecosystems under Climate Change and Ocean Acidification). As (M. Horsák et al. / Science of the Total Environment 634 (2018) 355–365) point out, some characteristics of ecosystems that make them especially vulnerable to climate change are (i) a preponderance of species that evolved under a cold climate, (ii) low productivity caused by nutrient limitation, which makes them sensitive to increased nutrient
cycling caused by warming (Cornelissen et al., 2007), and (iii) a scatterred distribution that limits species dispersal and migration (Pearson and Dawson, 2005). These characteristics are found in temperate peatlands such as mires, permanent semi-terrestrial wetlands where soils remain waterlogged but not inundated during most of the year (Wheeler and Proctor, 2000). Mires are priority habitats for conservation that harbour a high number of endangered species (Bergamini et al. 2009) and support highly-adapted flora in spatially reduced areas (Grootjans et al. 2006, Jiménez-Alfaro et al. 2013). Habitat distribution models have predicted a loss of mire surface in face of climate change (Essl et al. 2012).

Recent research has focused on the importance of groundwater as a regulator of mire microclimate. As azonal habitats whose occurence is driven by local edaphic conditions, locally obtained
thermal data are hence crucial for any prediction about the impact of climatically-induced warming (Breckle 2002). In Europe, fens are distributed from the Iberian Peninsula to Fennoscandia and from low valleys to the alpine belt, but nonetheless the constituent and dominant species remain relatively homogeneous (Jiménez-Alfaro et al. 2014). An apparent climatic resilience of fen species is further suggested by the persistence of relicts in their distribution limits (Amon et al. 2002, Topić and Stančić 2006, Hájek et al. 2009, Jiménez-Alfaro et al. 2012). A possible cause for this climatic resilience may be the buffer effect of groundwater on soil temperature, reducing the expected differences of the realized niche in diverse climatic regions (Ellenberg 1988). Indeed, the cool waters from underground aquifers that saturate fens are expected to keep temperatures warmer during cold periods and vice versa (Ellenberg 1988, Geiger et al. 2009). As root-zone temperature is known to be a major determinant of plant ecophysiology (Körner and Paulsen 2004), this buffer effect might allow fen species to regenerate and live along a wider range of air temperatures than they could otherwise. Soil temperature measurements have become available for mires of North America (Raney 2014), Western Europe (Fernández-Pascual et al. 2015), Central Europe (Horsak 2018). The comparison of said temperatures with air temperatures derived from models has shown that mire soil temperature is warmer in winter and cooler in summer, thus indicating the existence of a thermal buffer by groundwater (FerPas, Horsak). This effect has been linked to the composition of flora and fauna (Horsak 2018, Shecnkova 2020), growth of mire trees /Raney 2016) and the role of mires as glacial refugia (JimAlf 2016, Dite 2017). It is evident that the thermal buffer of groundwater will play a profound role in the response of these habitats to climate change, and needs to be studied in more depth.

>>>>

Climate change is already affecting global biodiversity patterns (IPCC 2014). Among other trends, there has been an expansion of drylands (Huang, J., Yu, H., Guan, X. et al. Accelerated dryland expansion under climate change), an increase of disturbations in forests (Seidl, R., Thom, D., Kautz, M. et al. Forest disturbances under climate change. Nature Clim Change 7, 395–402 (2017). https://doi.org/10.1038/nclimate3303) and ocean acidification (Coral Reef Ecosystems under Climate Change and Ocean Acidification). As (M. Horsák et al. / Science of the Total Environment 634 (2018) 355–365) point out, some characteristics of ecosystems that make them especially vulnerable to climate change are (i) a preponderance of species that evolved under a cold climate, (ii) low productivity caused by nutrient limitation, which makes them sensitive to increased nutrient
cycling caused by warming (Cornelissen et al., 2007), and (iii) a scatterred distribution that limits species dispersal and migration (Pearson and Dawson, 2005). These characteristics are found in temperate peatlands such as mires, permanent semi-terrestrial wetlands where soils remain waterlogged but not inundated during most of the year (Wheeler and Proctor, 2000). Mires are priority habitats for conservation that harbour a high number of endangered species (Bergamini et al. 2009) and support highly-adapted flora in spatially reduced areas (Grootjans et al. 2006). Habitat distribution models have predicted a loss of mire surface in face of climate change (Essl et al. 2012).

Recent research has focused on the importance of groundwater as a regulator of mire microclimate. As azonal habitats whose occurence is driven by local edaphic conditions, locally obtained
thermal data are hence crucial for any prediction about the impact of climatically-induced warming (Breckle 2002). In Europe, fens are distributed from the Iberian Peninsula to Fennoscandia and from low valleys to the alpine belt, but nonetheless the constituent and dominant species remain relatively homogeneous (Peterka). A possible cause for this climatic resilience may be the buffer effect of groundwater on soil temperature, reducing the expected differences of the realized niche in diverse climatic regions (Ellenberg 1988). Indeed, the cool waters from underground aquifers that saturate fens are expected to keep temperatures warmer during cold periods and vice versa (Ellenberg 1988, Geiger et al. 2009). As root-zone temperature is known to be a major determinant of plant ecophysiology (Körner and Paulsen 2004), this buffer effect might allow fen species to regenerate and live along a wider range of air temperatures than they could otherwise. Soil temperature measurements have become available for mires of North America (Raney 2014), Western Europe (Fernández-Pascual et al. 2015), Central Europe (Horsak 2018). The comparison of said temperatures with air temperatures derived from models has shown that mire soil temperature is warmer in winter and cooler in summer, thus indicating the existence of a thermal buffer by groundwater (FerPas, Horsak). This effect has been linked to the composition of flora and fauna (Horsak 2018, Shecnkova 2020), growth of mire trees /Raney 2016) and the role of mires as glacial refugia (JimAlf 2016, Dite 2017). It is evident that the thermal buffer of groundwater will play a profound role in the response of these habitats to climate change, and needs to be studied in more depth.
